---
title: |
  | Causal Inference for Observational Studies
  | Beyond the Basics
  | Class 1
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: Jake Bowers
bibliography:
 - ../BIB/references.bib
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: icpsr.beamer
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute
header-includes:
  - \setbeameroption{hide notes}
---

<!-- To show notes  -->
<!-- https://stackoverflow.com/questions/44906264/add-speaker-notes-to-beamer-presentations-using-rmarkdown -->

```{r echo=FALSE, include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
rm(list=ls())

require(knitr)

## This plus size="\\scriptsize" from https://stackoverflow.com/questions/26372138/beamer-presentation-rstudio-change-font-size-for-chunk
knitr::knit_hooks$set(mysize = function(before, options, envir) {
			      if (before)
				      return(options$size)
                              else return("\\normalsize")
})

knit_hooks$set(plotdefault = function(before, options, envir) {
		       if (before) par(mar = c(3, 3, .1, .1),oma=rep(0,4),mgp=c(1.5,.5,0))
})

opts_chunk$set(
	       tidy='styler',
	       echo=TRUE,
	       results='markup',
	       strip.white=TRUE,
	       fig.path='figs/fig',
	       cache=FALSE,
	       highlight=TRUE,
	       width.cutoff=132,
	       size='\\scriptsize',
	       out.width='.7\\textwidth',
	       fig.retina=FALSE,
	       message=FALSE,
	       comment=NA,
	       mysize=TRUE,
	       plotdefault=TRUE)

options(digits=4,
	scipen=8,
	width=132,
	show.signif.stars=FALSE)
```

```{r loadlibs,include=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(sandwich)
library(lmtest)
library(estimatr)
```

```{r makedirs, echo=FALSE, include=FALSE}
if(!dir.exists('figs')) dir.create('figs')

## Make a local library directory
if(!dir.exists(here::here('libraries'))){
        dir.create(here::here('libraries'))
}
```


```{r eval=TRUE, include=FALSE, echo=FALSE}
library('devtools')
library('withr')

with_libpaths(here::here('libraries'), install_github("markmfredrickson/RItools"), 'pre')
library('RItools',lib.loc = here::here('libraries'))
library(optmatch)
```

# Overview and Review

## Doing Matching, Assessing Designs, Estimating Effects

  1. **Review** Causal inference in randomized experiments; Statistical
     inference for causal effects in randomized experiments \autocite[Chap
     2]{rosenbaum2010}, \autocite[Chap 1-3]{gerbergreen2012}.

  2. **How to use optimal, full matching to produce a matched research
     design?**  Multivariate optimal matching review using `optmatch` in R:
     producing matched research designs using matching on scalars, propensity
     scores, and Mahalanobis distances \autocite{hansen2004}, \autocite[Chap
     1,3,7,8,9,13]{rosenbaum2010}, \autocite[Chap 9.0--9.2]{gelman2007dau}.

  3. **How to reason about whether we have a good matched research design?**
     Multivariate balance assessment using null hypothesis testing using the
     `RItools` package for R \autocite{hansenbowers2008} (and perhaps also
     using equivalence testing \autocite{hartman2018equivalence}).

  4. **How to customize and focus the matched research design creation?**
     Multivariate optimal matching; calipers; penalties; combining scores

  5. **How to estimate the ATE  and test hypotheses about causal effects given a
     matched research design?** Estimating ATE and Std Errors from Matched
     Designs using the Block Randomized Experiment as an Analogy.

## Overly Ambitious Plan

  - 09:00--09:30 ---  Introductions: Name, Affiliation/Organization, Interest (Substantive or Methodological)
  - 09:30--10:15 ---  Lecture by Jake to introduce concepts and encourage questions and answers.
  - 10:15--10:45 ---  Exercises by Class to raise even more questions.
  - 10:45 -- 11:00 --- Break
  - 11:00 -- 11:30 --- Lecture by Jake / question and answers about the previous lecture
  - 11:30 -- 12:00 --- Open Discussion on any topic.

Materials for practice can be downloaded from <http://jakeboweers.org/IPM2019>.

## What do you know about causal inference so far?

These words are used a lot in discussions of causal inference. Why are they important? What do they mean?

  - Potential Outcomes; Counterfactual Outcomes
  - Causal Effect
  - ATE
  - Randomization
  - Balance
  - Propensity Score
  - Mahalanobis Distance Score
  - Confounding; Selection Bias; Omitted Variable Bias
  - Covariates

What other words and/or phrases and/or concepts come to mind when you think about causal inference?

\medskip

Why is "randomization" listed here if a lot of the discussion about causal inference focuses on observational studies?

## Randomization and Causal Inference

From last week:

  - *Treatment* $Z_i=1$ for treatment and $Z_i=0$ for control for units $i$

  - Each unit has a pair of *potential outcomes* $(y_{i,Z_i=1},y_{i,Z_i=0})$
    (also written  $(y_{i,1},y_{i,0})$ ) (given SUTVA).

  - *Causal Effect*  when $y_{i,1} \ne y_{i,0}$, $\tau_i   =
    f(y_{i,1},y_{i,0})$ ex. $\tau_i =  y_{i,1} - y_{i,0}$.

  - *Fundamental Problem of (Counterfactual) Causality* We only see one
    potential outcome $Y_i = Z_i * y_{i,1} + (1-Z_i) y_{i,0}$

  - *Covariates*,  $\bX=\begin{bmatrix} x_{11} & \ldots & x_{1k} \\ \vdots &
    \vdots & \vdots \\  x_{n1} & \ldots & x_{nk} \end{bmatrix}$ is a matrix
    containing  background information about the units that might predict
    $(y_{i,1},y_{i,0})$ or $Z$ (except in an experiment).

Randomization makes $y_1, y_0, \bX \perp Z$. How to make use of this fact in a randomized experiment?

## Benefits of Randomized Designs

 1. Interpretable comparisons (lack of omitted variable bias, confounding,
    selection bias)
      - Can I interpret differences in outcome as caused by $Z$ and not $X$?
	Is it easy to confuse the effect of $Z$  with the  effects of $X$?
      - How does  randomization do this? How  does randomization eliminate
	**alternative explanations**?
 2. Reliable statistical inferences (estimators and tests)
      - The idea of **design-based** versus **model-based** statistical
	inference (next few slides).


## Design Based Approach 1: Test Hypotheses

 1. Make a guess about $\tau_i$.
 2. Then measure surprise or consistency of data with this guess given the
    design. (Given all of the ways this experiment could have  occurred, how
    many look more extreme than what we observe? Does our observation look
    typical or rare?)

\centering
  \includegraphics[width=.6\textwidth]{images/cartoonFisherNew1.pdf}

## Design Based Approach 1: Test Hypotheses

\centering
  \includegraphics[width=\textwidth]{images/cartoonFisherNew1.pdf}

## Design Based Approach 1: Test Hypotheses

\centering
  \includegraphics[width=\textwidth]{images/cartoonFisherNew2.pdf}



## Design Based Approach 2: Estimate Averages

  1. Notice that the observed $Y_i$ are a sample from  the (small, finite) population of $(y_{i,1},y_{i,0})$.
  2. Decide to focus on the average, $\bar{\tau}$, because sample averages, $\hat{\bar{\tau}}$ are unbiased and consistent estimators of population averages.
  3. Estimate $\bar{\tau}$ with the observed difference in means.

\centering
  \includegraphics[width=.5\textwidth]{images/cartoonNeyman.pdf}

## Design Based Approach 2: Estimate Averages

\centering
  \includegraphics[width=.9\textwidth]{images/cartoonNeyman.pdf}


## Approaches to creating interpretable comparisons:

   - Randomized experiments (more precision from reducing heterogeneity in $Y$)
   - Instrumental variables (with randomized $Z$ created $D$)
   - Natural Experiments / Discontinuities (one $X$ creates $Z$) (includes RDD)
   - Difference-in-Differences (reduce bias *and* increase precision from reducing heterogeneity in $Y$)
   - Semi-/Non-parametric Covariance Adjustment (ex. Matching)
   - Parametric covariance adjustment

## Lingering Questions?


# Doing Multivariate Matching

```{r loaddata, echo=FALSE, cache=TRUE}
load(url("http://jakebowers.org/Data/meddat.rda"))
row.names(meddat) <- meddat$nh ## add row.names for use later
```

##  Did the Metrocable Reduce Crime?

Cerdá et al. collected data on about roughly `r nrow(meddat)`
neighborhoods in Medellin, Colombia. About  `r signif(sum(meddat$nhTrt),2)` of had
access to the new Metrocable line and `r signif(sum(1-meddat$nhTrt),2)` did not.

\centering
\includegraphics[width=.7\textwidth]{medellin-gondola.jpg}

<!-- For more on the Metrocable project see <https://www.medellincolombia.co/where-to-stay-in-medellin/medellin-orientation/> and <https://archleague.org/article/connective-spaces-and-social-capital-in-medellin-by-jeff-geisinger/> -->



##  Did Transportation Reduce Crime?

Cerdá et al. collected data on about roughly `r nrow(meddat)`
neighborhoods in Medellin, Colombia. About  `r signif(sum(meddat$nhTrt),2)` of had
access to the new Metrocable line and `r signif(sum(1-meddat$nhTrt),2)` did not.

\centering
\includegraphics[width=.8\textwidth]{medellin-conc-pov.jpg}


## Codebook/Covariates

We don't have a codebook. Here are some guesses about only some of the variables.
listed here.

\scriptsize
```
## The Intervention
nhTrt        Intervention neighborhood (0=no Metrocable station, 1=Metrocable station)

## Some Covariates (there are others, see the paper itself)
nh03         Neighborhood id
nhGroup      Treatment (T) or Control (C)
nhTrt        Treatment (1) or Control (0)
nhHom        Mean homicide rate per 100,000 population in 2003
nhDistCenter Distance to city center (km)
nhLogHom     Log Homicide (i.e. log(nhHom))

## Outcomes (BE03,CE03,PV03,QP03,TP03 are baseline versions)
BE      Neighborhood amenities Score 2008
CE      Collective Efficacy Score 2008
PV      Perceived Violence Score 2008
QP      Trust in local agencies Score 2008
TP      Reliance on police Score 2008
hom     Homicide rate per 100,000 population Score 2008-2003 (in log odds)

HomCount2003 Number of homicides in 2003
Pop2003      Population in 2003
HomCount2008 Number of homicides in 2008
Pop2008      Population in 2008
```

## How did the Metrocable effect Homicides?


```{r ratesfromcounts, echo=FALSE}
meddat<- mutate(meddat, HomRate03=(HomCount2003/Pop2003)*1000,
                HomRate08=(HomCount2008/Pop2008)*1000)
row.names(meddat) <- meddat$nh ##tidyverse doesn't like row.names
```

One approach:  Estimate the average treatment effect of Metrocable on
Homicides after the stations were built.

```{r threeestates}
themeans<-group_by(meddat,nhTrt) %>% summarise(ybar=mean(HomRate08))
themeans
diff(themeans$ybar)
lmOne <- lm(HomRate08~nhTrt,meddat)
coef(lmOne)["nhTrt"]
library(estimatr)
difference_in_means(HomRate08~nhTrt,meddat)
```

\bigskip

**But alternative explanations?**

## What are alternative explanations for this effect?

We claim that the policy intervention had some effect. What are some alternative explanations?

## Evidence about confounding?

The `xBalance` function in the `RItools` R package tests the hypothesis that treatment assignment is independent of covariates \citep{bowersfredericksonhansen2016}.

```{r}
## xBalance(treatment~covariate)
xbMed1 <- xBalance(nhTrt~nhAboveHS,data=meddat,report="all")
xbMed1$overall ##  the overall chi-sq test
xbMed1$results[1,c("Control","Treatment","adj.diff","std.diff","z","p"),]  ## the covariate specific z-test
```

It would be rare/surprising/strange to see a Treatment-vs-Control difference in
education if treatment were independent of education.

\medskip

\emph{Note on `xBalance`}: Treatment  would be independent of education if it
were randomly assigned. So, `xBalance` compares the observed (standardized)
mean difference  to those consistent with a randomized experiment.

## How would you adjust for Proportion Above HS Degree?

So, part of the Metrocable effect might not reflect the causal effect of
Metrocable per se, but rather the education of people in the
neighborhood. How should we remove `nhAboveHS` from our estimate or test? What
strategies can you think of?

## Stratification V 1.0

```{r}
lm1a <- lm(HomRate08~nhTrt,data=meddat,subset=nhAboveHS>=.1)
lm1b <- lm(HomRate08~nhTrt,data=meddat,subset=nhAboveHS<.1)
res_strat <- c(hiEd_Effect=coef(lm1a)["nhTrt"],loEd_Effect= coef(lm1b)["nhTrt"])
res_strat
n_strat <- table(meddat$nhAboveHS>=.1)
n_strat
stopifnot(sum(n_strat)==nrow(meddat)) ## A test of code
sum(res_strat * rev(n_strat)/45) ## What is happening here?
```

*But, standard errors? p-values? confidence intervals?*

## Stratified adjustment V 2.0

One-step stratified estimation.

```{r}
## Weight by block size
ate1c <- difference_in_means(HomRate08~nhTrt, blocks = I(nhAboveHS>=.1),data=meddat)
ate1c
## Weight by both block size and  proportion in treatment vs control ("harmonic weight")
lm1c <- lm_robust(HomRate08~nhTrt, fixed_effects = ~I(nhAboveHS>=.1),data=meddat)
coef(lm1c)["nhTrt"]
lm1d <- lm(HomRate08~nhTrt+I(nhAboveHS>=.1),data=meddat)
coef(lm1d)["nhTrt"]
xbate1 <- xBalance(nhTrt~HomRate08,strata=list(hs=~I(nhAboveHS>=.1)),data=meddat,report="all")
xbate1$results[1,c("Control","Treatment","adj.diff"),]
```

## Balance assessment after stratification

Did we adjust enough? What would *enough* mean?

```{r}
xbHS1 <- xBalance(nhTrt~nhAboveHS,strata=list(hs=~I(nhAboveHS>=.1)),data=meddat,report="all")
xbHS1$overall
xbHS1$results[1,c("Control","Treatment","adj.diff","std.diff","z","p"),]  ## the covariate specific z-test
```
## Disadvantages and Advantages of Simple Stratification

 -  (+) Easy to explain what  "controlling for" or "adjustment" means.
 -  (-) Hard to justify any particular cut-point
 -  (-) We could probably adjust *more* --- comparing neighborhoods similar in education rather than just  within  big   strata

## Optmatch workflow: Stratification by minimizing distances

Introduction to `optmatch` workflow. To minimize differences requires a matrix
of those differences (in general terms, a matrix of distances between the
treated and control units)

```{r}
## Using the help page for match_on for scalar distances
tmp <- meddat$nhAboveHS
names(tmp) <- rownames(meddat)
absdist <- match_on(tmp, z = meddat$nhTrt,data=meddat)
absdist[1:3,1:10]
abs(meddat$nhAboveHS[meddat$nhTrt==1][1] - meddat$nhAboveHS[meddat$nhTrt==0][1] )
```

## Create Matched Research Designs

```{r}
fm1 <- fullmatch(absdist,data=meddat)
summary(fm1, min.controls=0, max.controls=Inf )
table(meddat$nhTrt,fm1)

pm1 <- pairmatch(absdist,data=meddat)
summary(pm1, min.controls=0, max.controls=Inf )
table(meddat$nhTrt,pm1,exclude=c())
```

## Evaluate the design: Within set differences

```{r echo=FALSE}
meddat$fm1 <- fm1
meddat$pm1 <- pm1
```

Differences within sets versus raw differences.

```{r echo=FALSE, out.width=".9\\textwidth"}
library(gridExtra)
bpfm1 <- ggplot(meddat,aes(x=fm1,y=nhAboveHS)) +
	geom_boxplot() +
	 stat_summary(fun.y=mean, geom="point", shape=20, size=3, color="red", fill="red")
meddat$nostrata <- rep(1,45)
bporig <- ggplot(meddat,aes(x=nostrata,y=nhAboveHS))+
	 geom_boxplot()+
	 stat_summary(fun.y=mean, geom="point",
		      shape=20, size=3, color="red", fill="red")

grid.arrange(bpfm1,bporig,ncol=2,layout_matrix=matrix(c(1,1,1,1,2),nrow=1))
```

## Evaluate the design: Within set differences

```{r echo=FALSE, out.width=".9\\textwidth"}
bppm1 <- ggplot(meddat,aes(x=pm1,y=nhAboveHS)) +
	geom_boxplot() +
	 stat_summary(fun.y=mean, geom="point", shape=20, size=3, color="red", fill="red")

grid.arrange(bppm1,bporig,ncol=2,layout_matrix=matrix(c(1,1,1,1,2),nrow=1))
```

## Evaluate the design: Inspecet within set differences

```{r echo=FALSE}
rawmndiffs <- with(meddat, mean(nhAboveHS[nhTrt==1]) - mean(nhAboveHS[nhTrt==0]))
setdiffsfm1 <- meddat %>% group_by(fm1) %>% summarize(mneddiffs =
						   mean(nhAboveHS[nhTrt==1]) -
						   mean(nhAboveHS[nhTrt==0]),
					   mnAboveHS = mean(nhAboveHS),
					   minAboveHS = min(nhAboveHS),
					   maxAboveHS = max(nhAboveHS))

setdiffsfm1
#summary(setdiffs$mneddiffs)
#quantile(setdiffs$mneddiffs, seq(0,1,.1))
```

## Evaluate the design: Inspect within set differences


```{r echo=FALSE, warnings=FALSE}
setdiffspm1 <- meddat %>% group_by(pm1) %>% summarize(mneddiffs =
						   mean(nhAboveHS[nhTrt==1]) -
						   mean(nhAboveHS[nhTrt==0]),
					   mnAboveHS = mean(nhAboveHS),
					   minAboveHS = min(nhAboveHS),
					   maxAboveHS = max(nhAboveHS))

setdiffspm1
```

## Evaluate the design: Compare to a randomized experiment.

The within-set differences look different from those that would be expected
from a randomized experiment.

```{r echo=FALSE}
xbHS2 <- xBalance(nhTrt~nhAboveHS,
                  strata=list(nostrat=NULL,
                              fm = ~fm1,
			      pm = ~pm1),
                  data=meddat,report="all")
xbHS2$results
xbHS2$overall
```


## What is xBalance doing?

```{r}
setmeanDiffs <- meddat %>% group_by(fm1) %>%
  summarise(diffAboveHS=mean(nhAboveHS[nhTrt==1])-mean(nhAboveHS[nhTrt==0]),
            nb=n(),
            nTb = sum(nhTrt),
            nCb = sum(1-nhTrt),
            hwt = ( 2*( nCb * nTb ) / (nTb + nCb))
            )
setmeanDiffs
```

## What is xBalance doing with multiple sets/blocks?

The test statistic is a weighted average of the set-specific differences (same
approach as we would use to test the null in a block-randomized experiment)

```{r}
## The descriptive adj.mean diff from balanceTest
with(setmeanDiffs, sum(diffAboveHS*nTb/sum(nTb)))
## The mean diff used as the observed value in the testing
with(setmeanDiffs, sum(diffAboveHS*hwt/sum(hwt)))
## Compare to xBalance output
xbHS2$results[,,"fm"]
```

Notice that `balanceTest` prints the set-size weighted difference (the updated
version differs a little from `xBalance`):

```{r}
btHS2 <- balanceTest(nhTrt~nhAboveHS+strata(fm1) +strata(pm1),
                  data=meddat,report="all")
btHS2
```

#  Matching With Many Covariates: Using Mahalnobis Distance

## Dimension reduction using the Mahalanobis Distance

The general idea: dimension reduction. When we convert many columns into one
column we reduce the dimensions of the dataset (to one column).


```{r}
X <- meddat[,c("nhAboveHS","nhPopD")]
row.names(X) <- meddat$nh
plot(meddat$nhAboveHS,meddat$nhPopD,xlim=c(-.3,.6),ylim=c(50,700))
```

## Dimension reduction using the Mahalanobis Distance

First, let's look at Euclidean distance: $\sqrt{ (x_1 - x_2)^2 + (y_1 - y_2)^2 }$

```{r echo=FALSE, out.width=".8\\textwidth"}
par(mgp=c(1.25,.5,0),oma=rep(0,4),mar=c(3,3,0,0))
plot(meddat$nhAboveHS,meddat$nhPopD,xlim=c(-.3,.6),ylim=c(50,700))
points(mean(X[,1]),mean(X[,2]),pch=19,cex=1)
arrows(mean(X[,1]),mean(X[,2]),X["407",1],X["407",2])
text(.4,200,label=round(dist(rbind(colMeans(X),X["407",])),2))
```

## Dimension reduction using the Mahalanobis Distance

First, let's look at Euclidean distance: $\sqrt{ (x_1 - x_2)^2 + (y_1 - y_2)^2 }$

```{r echo=FALSE, out.width=".5\\textwidth"}
par(mgp=c(1.25,.5,0),oma=rep(0,4),mar=c(3,3,0,0))
plot(meddat$nhAboveHS,meddat$nhPopD,xlim=c(-.3,.6),ylim=c(50,700))
points(mean(X[,1]),mean(X[,2]),pch=19,cex=1)
arrows(mean(X[,1]),mean(X[,2]),X["407",1],X["407",2])
text(.4,200,label=round(dist(rbind(colMeans(X),X["407",])),2))
```

```{r}
tmp <- rbind(colMeans(X),X["407",])
tmp
sqrt( (tmp[1,1] - tmp[2,1])^2 + (tmp[1,2]-tmp[2,2])^2 )
```

## Dimension reduction using the Mahalanobis Distance

Now the Euclidean distance (on a standardized scale)

```{r echo=FALSE}
Xsd <-scale(X)
apply(Xsd,2,sd)
apply(Xsd,2,mean)
plot(Xsd[,1],Xsd[,2],xlab="nhAboveHS/sd",ylab="nhPopD/sd")
points(mean(Xsd[,1]),mean(Xsd[,2]),pch=19,cex=1)
arrows(mean(Xsd[,1]),mean(Xsd[,2]),Xsd["407",1],Xsd["407",2])
text(2,-1.2,label=round(dist(rbind(colMeans(Xsd),Xsd["407",])),2))
```


## Dimension reduction using the Mahalanobis Distance

The Mahalanobis distance avoids the scale problem in the euclidean distance.
<!-- [For more see <https://stats.stackexchange.com/questions/62092/bottom-to-top-explanation-of-the-mahalanobis-distance>] -->

```{r echo=FALSE, out.width=".6\\textwidth"}
library(chemometrics)
par(mgp=c(1.5,.5,0),oma=rep(0,4),mar=c(3,3,0,0))
mh <- mahalanobis(X,center=colMeans(X),cov=cov(X))
drawMahal(X,center=colMeans(X),covariance=cov(X),
          quantile = c(0.975, 0.75, 0.5, 0.25))
abline(v=mean(meddat$nhAboveHS),h=mean(meddat$nhPopD))
pts <-c("401","407","411","202")
arrows(rep(mean(X[,1]),4),rep(mean(X[,2]),4),X[pts,1],X[pts,2])
text(X[pts,1],X[pts,2],labels=round(mh[pts],2),pos=1)
```

```{r}
Xsd <- scale(X)
tmp<-rbind(c(0,0),Xsd["407",])
mahalanobis(tmp,center=c(0,0),cov=cov(Xsd))
edist <- sqrt( (tmp[1,1] - tmp[2,1])^2 + (tmp[1,2]-tmp[2,2])^2 )
edist
```

## Matching on the Mahalanobis Distance

Here using the rank based Mahalanobis distance following DOS Chap. 8 (but
comparing to the ordinary version).

```{r}
mhdist <- match_on(nhTrt~nhPopD+nhAboveHS,data=meddat,method="rank_mahalanobis")
mhdist[1:3,1:8]
mhdist2 <- match_on(nhTrt~nhPopD+nhAboveHS,data=meddat)
mhdist2[1:3,1:8]
```

## A Mahalanobis Distance Picture

```{r echo=FALSE}
cpts <-c("401","407","411")
tpts <-c("101","102","202")
```

Distances between neighborhood 407 and some others.
```{r}
mhdist2[tpts,"407"]
```

```{r distto407, echo=FALSE}
par(mgp=c(1.5,.5,0),oma=rep(0,4),mar=c(3,3,0,0))
drawMahal(X,center=colMeans(X),covariance=cov(X),
          quantile = c(0.975, 0.75, 0.5, 0.25))
abline(v=mean(meddat$nhAboveHS),h=mean(meddat$nhPopD))
arrows(X[tpts,1],X[tpts,2],rep(X["407",1]),rep(X["407",2]))
text(X[tpts,1],X[tpts,2],labels=round(mhdist2[tpts,"407"],2),pos=1)
```


## Matching on the Mahalanobis Distance

```{r}
fmMh <- fullmatch(mhdist,data=meddat)
summary(fmMh,min.controls=0,max.controls=Inf)
```

#  Matching on Many Covariates: Using Propensity Scores

## Matching on the propensity score

**Make the score**^[Note that we will be using `brglm` or `bayesglm` in the
future because of logit separation problems when the number of covariates
increases.]

```{r}
theglm <- glm(nhTrt~nhPopD+nhAboveHS,data=meddat,family=binomial(link="logit"))
thepscore <- theglm$linear.predictor
thepscore01 <- predict(theglm,type="response")
````

We tend to match on the linear predictor rather than the version required to
range only between 0 and 1.

```{r echo=FALSE, out.width=".7\\textwidth"}
par(mfrow=c(1,2),oma=rep(0,4),mar=c(3,3,2,0),mgp=c(1.5,.5,0))
boxplot(split(thepscore,meddat$nhTrt),main="Linear Predictor (XB)")
stripchart(split(thepscore,meddat$nhTrt),add=TRUE,vertical=TRUE)

boxplot(split(thepscore01,meddat$nhTrt),main="Inverse Link Function (g^-1(XB)")
stripchart(split(thepscore01,meddat$nhTrt),add=TRUE,vertical=TRUE)
```

## Matching on the propensity score

```{r}
psdist <- match_on(theglm,data=meddat)
psdist[1:4,1:4]
fmPs <- fullmatch(psdist,data=meddat)
summary(fmPs,min.controls=0,max.controls=Inf)
```

## Can you do better?

**Challenge:** Improve the matched design by adding covariates or functions of
covariates using either or both of the propensity score or mahalanobis distance
(rank- or not-rank based). So far we have:

```{r}
thecovs <- unique(c(names(meddat)[c(5:7,9:24)],"HomRate03"))
balfmla<-reformulate(thecovs,response="nhTrt")
xb4 <- balanceTest(update(balfmla,.~.+strata(fmMh)+strata(fmPs)),
                   data=meddat,report="all",p.adjust.method="none")
xb4$overall[,]
```

## Can you do better?

Challenge: Improve the matched design. So far we have:

```{r}
plot(xb4)
```

# Estimating Treatment Effects and Testing Hypotheses About Causal Effects

## The As-If-Randomized Approach to Estimating the ATE

A simple approach:

```{r}

meddat$fmMh <-  fmMh

ateEst1 <- difference_in_means(HomRate08~nhTrt,blocks=fmMh,data=meddat)
ateEst2 <- lm_robust(HomRate08~nhTrt,fixed_effects=~fmMh,data=meddat)

```

What is happening here? Why do they differ?


# Summary of the day

## What have we done?

 - Reviewed introductory material
 - Discussed a method to compare observed data to a randomized design as a way to answer the question about "enough adjustment"
 - Used simple stratification for adjustment, and assessed the performance of that approach.
 - Created matched  designs using scalar as well as multivariate distances.
 - Estimated the ATE and tested the hypothesis of no effects given the matched research designs that we created.

# Supplementary Material

## One approach to this problem: model-based adjustment

Let's try to just adjust for this covariate in a very common manner:

```{r echo=FALSE}
lm1 <- lm(HomRate08~nhTrt+nhAboveHS,data=meddat)
```

```{r echo=FALSE}
preddat <- expand.grid(nhTrt=c(0,1),nhAboveHS=range(meddat$nhAboveHS))
preddat$fit <- predict(lm1,newdata=preddat)
```

\centering
```{r, out.width=".9\\textwidth", echo=FALSE}
par(oma=rep(0,4),mgp=c(1.5,.5,0),mar=c(3,3,0,0))
with(meddat, plot(nhAboveHS,HomRate08,pch=c(1,2)[nhTrt+1]))
with(subset(preddat,subset=nhTrt==0),lines(nhAboveHS,fit,lty=1))
with(subset(preddat,subset=nhTrt==1),lines(nhAboveHS,fit,lty=2))
## locator()
text(c(0.111807,0.001629), c(1.871,2.204), labels=c("Treat","Control"),pos=1)
text(c(.3,.5),c( coef(lm1)[1]+coef(lm1)[3]*.3 , coef(lm1)[1]+coef(lm1)[2]+coef(lm1)[3]*.5),
     labels=c("Control","Treat"))
```

## Exactly what does this kind of adjustment do?

Notice that I can get the same coefficient (the effect of Metrocable on
Homicides adjusted for HS-Education in the neighborhood) either directly (as
earlier) or via **residualization**:

```{r}
coef(lm1)["nhTrt"]
eYX <- residuals(lm(HomRate08~nhAboveHS,data=meddat))
eZX <- residuals(lm(nhTrt ~ nhAboveHS, data=meddat))
lm1a <- lm(eYX~eZX)
coef(lm1a)[2]
```

## Exactly what does this kind of adjustment do?

So, how would you explain what it means to "control for HS-Education" here?

```{r}
plot(eZX,eYX)
```



## Did we adjust enough?

Maybe adding some more information to the plot can help us decide whether, and to what extend, we effectively "controlled for" the proportion of the neighborhood with more than High School education. Specifically, we might be interested in assessing extrapolation/interpolation problems arising from our linear assumptions.

\centering
```{r, out.width=".7\\textwidth", echo=FALSE, warning=FALSE, message=FALSE}
par(oma=rep(0,4),mgp=c(1.5,.5,0),mar=c(3,3,0,0))
with(meddat, plot(nhAboveHS,HomRate08,pch=c(1,2)[nhTrt+1]))
with(subset(preddat,subset=nhTrt==0), lines(nhAboveHS,fit,lty=1))
with(subset(preddat,subset=nhTrt==1),lines(nhAboveHS,fit,lty=2))
with(subset(meddat,subset=nhTrt==0),lines(loess.smooth(nhAboveHS,HomRate08,deg=1,span=2/3),lty=1))
with(subset(meddat,subset=nhTrt==1),lines(loess.smooth(nhAboveHS,HomRate08,deg=1,span=2/3),lty=2))
## locator()
text(c(0.111807,0.001629), c(1.871,2.204), labels=c("Treat","Control"),pos=1)
text(c(.3,.5),c( coef(lm1)[1]+coef(lm1)[3]*.3 , coef(lm1)[1]+coef(lm1)[2]+coef(lm1)[3]*.5),
     labels=c("Control","Treat"))
with(subset(meddat,subset=nhTrt==0),rug(nhAboveHS))
with(subset(meddat,subset=nhTrt==1),rug(nhAboveHS,line=-.5))
```

How should we interpret this adjustment? How should we judge the improvement that we made? What concerns might we have?



```{r echo=FALSE, eval=FALSE}
thecovs <- unique(c(names(meddat)[c(5:7,9:24)],"HomRate03"))
balfmla<-reformulate(thecovs,response="nhTrt")

xbMed<-balanceTest(balfmla,
	      data=meddat,
	      report=c("all"),
        p.adjust.method="none")
xbMed$overall
xbMed$results["nhAboveHS",,]
```


```{r echo=FALSE, eval=FALSE}
outcomefmla <- reformulate(c("nhTrt",thecovs),response="HomRate08")
lmbig <- lm(outcomefmla,data=meddat)
```

## Back to the randomized experiment to help make a case for appropriate adjustment

It seems like arguments about plots may be difficult to resolve. How about going back to `xBalance`? The idea of a standard against which we can compare a given design. How might we do this in this case?


*Attempt 1: Stratification ... but two estimates rather than one adjusted
estimate*

```{r}
lm1a <- lm(HomRate08~nhTrt,data=meddat,subset=nhAboveHS>=.1)
lm1b <- lm(HomRate08~nhTrt,data=meddat,subset=nhAboveHS<=.1)
```

*Attempt 2: Stratified adjustment*

```{r}
lm1c <- lm(HomRate08~nhTrt+I(nhAboveHS>=.1),data=meddat)
coef(lm1c)["nhTrt"]
```

And balance assessment after stratification:

```{r}
xbHS1 <- xBalance(nhTrt~nhAboveHS,strata=list(hs=~I(nhAboveHS>=.1)),data=meddat,report="all")
xbHS1$overall
xbHS1$results
```

## The Curse of Dimensionality and linear adjustment for one more variable.

What about more than one variable? Have we controlled for both population
density and educational attainment enough? How would we know?

```{r}
lm2x <- lm(HomRate08 ~ nhTrt + nhPopD + nhAboveHS, data=meddat)
coef(lm2x)["nhTrt"]
```

Maybe another plot?

```{r eval=FALSE}
meddat$nhTrtF <- factor(meddat$nhTrt)
library(car)
scatter3d(HomRate08~nhAboveHS+nhPopD,
	  groups=meddat$nhTrtF,
	  data=meddat,surface=TRUE,
    fit=c("linear")) #additive"))
```

```{r echo=FALSE, eval=FALSE}
scatter3d(HomRate08~nhAboveHS+nhPopD,
	  groups=meddat$nhTrtF,
	  data=meddat,surface=TRUE,
    fit=c("additive"))

```


## Can we improve stratified adjustment?

Rather than two strata, why not three?

```{r}
lm1cut3 <- lm(HomRate08~nhTrt+cut(nhAboveHS,3),data=meddat)
coef(lm1cut3)["nhTrt"]
```
But why those cuts? And why not 4? Why not...?

\medskip

One idea: collect observations into strata such that the sum of the
differences in means of nhAboveHS within strata is smallest? This is the idea
behind `optmatch` and other matching approaches.

## References

