
## What about controlling for x1?

If we knew that x1 causes both Z and Y, then perhaps we can "control for it"?
What does this do? Say we do not know how Z was assigned. Which should we use?

```{r echo=TRUE}
set.seed(12345)
dat$Z <- new_exp2(dat$Z,dat$x1)
e1 <-lm(Y~Z,data=dat)
e2 <- lm(Y~Z+x1,data=dat)
coef(e1)[["Z"]]
coef(e2)[["Z"]]
```

## What about controlling for x1?

What does it mean when we "control for" x1?

```{r}
with(dat,plot(x1,Y,pch=c(1,19)[Z+1],col=c("black","grey")[Z+1]))
abline(a=coef(e2)[[1]],b=coef(e2)[["x1"]])
abline(a=coef(e2)[[1]]+1,b=coef(e2)[["x1"]])
text(y=c(coef(e2)[[1]],coef(e2)[[1]]+1),x=c(0,0),labels=c("Z=0","Z=1"))
```

## What about controlling for x1?

Controlling for as residualization:

```{r echo=TRUE}
resid1 <- residuals(lm(Y~x1,data=dat))
resid2 <- residuals(lm(Z~x1,data=dat))
e2new <- lm(resid1~resid2)
coef(e2new)[[2]]
stopifnot(all.equal(coef(e2new)[[2]], coef(e2)[["Z"]]))
```

```{r out.width=".5\\textwidth"}
par(mfrow=c(1,2),oma=rep(0,4),mar=c(3,3,1,1),mgp=c(1.25,.5,0))
with(dat,plot(x1,Y))
abline(lm(Y~x1,data=dat))
plot(dat$x1,resid1)
abline(lm(resid1~dat$x1))
```


## What about controlling for x1?

Controlling for as residualization:

```{r out.width=".8\\textwidth"}
par(mfrow=c(1,2),oma=rep(0,4),mar=c(3,3,1,1),mgp=c(1.25,.5,0))
with(dat,plot(x1,Y))
abline(lm(Y~x1,data=dat))
plot(dat$x1,resid1)
abline(lm(resid1~dat$x1))
```

## What about controlling for x1?

**How** should we control for x1? How to choose?

```{r echo=TRUE}
set.seed(12345)
dat$Z <- new_exp2(dat$Z,dat$x1)
e1 <-lm(Y~Z,data=dat)
e2 <- lm(Y~Z+x1,data=dat)
e3 <- lm(Y~Z+x1+I(x1^2),data=dat)
e4 <- lm(Y~Z*x1+I(x1^2),data=dat)
coef(e1)[["Z"]]
coef(e2)[["Z"]]
coef(e3)[["Z"]]
preddat <- expand.grid(Z=unique(dat$Z),x1=unique(dat$x1))
preddat$e4fit <- predict(e4,newdata=preddat)
## Equal weight estimate
with(preddat,mean(e4fit[Z==1]) - mean(e4fit[Z==0]))
```

Is this enough? What else might we do? Did we do a good job? How would we know? What does "controlling for" really mean?


