---
title: 'Exercise 1: Describe your data and the causal inference problem'
author: 'Jake Bowers'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
header-includes:
  - \usepackage{bm}
output:
  pdf_document:
    number_sections: true
    fig_caption: yes
    fig_height: 8
    fig_width: 8
    latex_engine: xelatex
    citation_package: biblatex
    keep_tex: true
fontsize: 10pt
geometry: margin=1in
graphics: yes
mainfont: "Helvetica"
bibliography: classbib.bib
biblio-style: "authoryear-comp,natbib"
---

<!-- Make this document using library(rmarkdown); render("exploration1.Rmd") -->
<!-- \input{mytexsymbols} -->

```{r setup, echo=FALSE, results=FALSE, include=FALSE, cache=FALSE}
library(here)
source(here("rmd_setup.R"))
```

This exercise should help you (1) get some practice with R, (2) find some data
that you can use to learn about matching and stratification for causal
inference and adjustment, and (3) maybe even make some progress on some project.

None of these exercises are graded. These questions are just meant to help you
made progress.

I will break down the questions into very basic steps (perhaps even too basic)
in order to help with clarity.

1. What is the causal relationship that you want to learn about? (For example,
   using the data example used in the class, "What is the effect of building
   the Metrocable stations on violence in the neighborhoods of Medellin?")

2. What is the **outcome** that you want to learn about? (For example,
   "Violence in the neighborhoods of Medellin. Especially, the murder rate")
   Please show a description of the outcome here. It could be a table or a
   graph. The units should be clear.

3. What is the **intervention** or **treatment** you are studying? (For
   example, "The building of a Metrocable station --- which is a bundle of
   public policy interventions that improve the area around the new
   transpotation station, including public spaces and police presence, etc.")
   Please show a description of the intervention or outcome. This can be binary
   (for use with **non-bipartite matching** or linear models) or categorial or continuous (for
   us with **non-bipartite** matching or linear models).

4. What **background covariates** do you worry might confuse interpretation of the simple bivariate relationship between intervention (or cause) and outcome (or effect)? (For example, "I worry that stations might have been built in places with high violence rates. So comparing violence rates between neighborhoods with stations and neighborhoods without stations might show higher violence in places with stations --- even if the stations caused a reduction in violence.")  Please show some descriptions of the covariates here.


5. How do these covariates relate to the intervetion or treatment? Is there
   reason to worry that much about such confounding?


Note: If you do not have your own data, feel free to use the Medellin data:

> In 2004 the municipality of Medell\'{i}n, Columbia built the first line
 of the Metrocable --- a set of cable cars that connected poor neighborhoods
 on the edges of the city to the center of the city \citep{cerda2012reducing}.
 Professor Magdalena Cerd\'{a} and her collaborators asked whether this kind
 of integration could improve life in these poor (and heretofore violent)
 neighborhoods. We ~~extracted~~ were given some of the data from this project to use
 here.\footnote{The articles can be both found in this web directory
 \url{http://jakebowers.org/Matching/}.}

```{r}
load(url("http://jakebowers.org/Data/meddat.rda"))
```

> We don't have a formal codebook. Here are some guesses about the meanings of
some of the variables. There are more variables in the data file than those
listed here.

```
## The Intervention
nhTrt        Intervention neighborhood (0=no Metrocable station, 1=Metrocable station)

## Some Covariates (there are others, see the paper itself)
nh03         Neighborhood id
nhGroup      Treatment (T) or Control (C)
nhTrt        Treatment (1) or Control (0)
nhHom        Mean homicide rate per 100,000 population in 2003
nhDistCenter Distance to city center (km)
nhLogHom     Log Homicide (i.e. log(nhHom))

## Outcomes (BE03,CE03,PV03,QP03,TP03 are baseline versions)
BE      Neighborhood amenities Score 2008
CE      Collective Efficacy Score 2008
PV      Perceived Violence Score 2008
QP      Trust in local agencies Score 2008
TP      Reliance on police Score 2008
hom     Homicide rate per 100,000 population Score 2008-2003 (in log odds)

HomCount2003 Number of homicides in 2003
Pop2003      Population in 2003
HomCount2008 Number of homicides in 2008
Pop2008      Population in 2008
```

# References

